Two years of explicit CiTO annotations

### 2023-01-08 article.Rmd
Added metadata
Revised the document, clarifying details

---
title: "Two years of explicit CiTO annotations"
subtitle: "Type: Correspondence"
author:
- Egon Willighagen [1,*,orcid:0000-0001-7542-0286]
link-citations: yes
geometry: margin=2.5cm
output:
  html_document:
    pandoc_args:
    - --lua-filter=filters/extract-cito.lua
    - --citeproc
    - --lua-filter=filters/insert-cito-in-ref.lua
  pdf_document:
    pandoc_args:
    - --lua-filter=filters/extract-cito.lua
    - --citeproc
    - --lua-filter=filters/insert-cito-in-ref.lua
  word_document:
    pandoc_args:
    - --lua-filter=filters/extract-cito.lua
    - --citeproc
    - --lua-filter=filters/insert-cito-in-ref.lua
# RStudio: cito-bibliography: bibliography.bib
# Other: bibliography: bibliography.bib
cito-bibliography: bibliography.bib
csl: journal-of-cheminformatics.csl
...

@@

@@

Communicating new research findings is primarily done by written texts in the form of scholarly
articles, books, and book chapters. To not having to repeat past research by themselves or others,
authors cite research relevant [@agreesWith:citesAsAuthority:Shotton_2010]. However, the reasons why
authors cite literature vary, which complicates how we use citations [@agreesWith:citesAsAuthority:Nicholson_2021]. 
Typing citations is therefore of interest: it allows us to navigate literature more easily: it points
us to essential research methods, data, and can warn us of research that cannot be reproduced,
or others disagree with [@Nicholson_2021].

@@

The pilot consists of a couple of components and the editorial explains some of them [@Willighagen_2020].
The Citation Typing Ontology was selected to express the intention [@usesMethodIn:Shotton_2010],
the intention is expressed a compact identifier wrapped in square brackets, also called a
safe CURIE, standard proposed by the W3C in 2010 [@usesMethodIn:McCarron2010,@usesMethodIn:Wimalaratne_2018].
The *cito* prefix is registered in Bioregistry [@usesDataFrom:Hoyt2022].
The *bibnotes* concept of the Springer Nature publishing platform was used as carrier.
Authors are guided by a landing page consisting of a BMC Collection at
[https://www.biomedcentral.com/collections/cito](https://www.biomedcentral.com/collections/cito)
and author guidelines explaining to authors
how they can add the annotations with their favorite editor at
[https://jcheminform.github.io/jcheminform-author-guidelines/cito](https://jcheminform.github.io/jcheminform-author-guidelines/cito).

Because the CiTO ontology has many terms for many different citation intentions,
we made a selection of CiTO terms authors could use [@CiTOCollection]:
**[cito:citesAsDatasource]** to indicate
to a source the provides data to back up a claim, **[cito:usesDataFrom]** to indicate that the authors
reused data, **[cito:usesMethodIn]** when a method or protocol explain in that source is used,
and a few more general intentions like **[cito:discusses]**, **[cito:extends]**, **[cito:agreesWith]**,
and **[cito:disagreesWith]**.
The journal itself would adopt the following CiTO annotations: **[cito:retracts]**,
**[cito:repliesTo]**, and **[updates]**. Fortunately, it has not been used yet, but
the first would be used if an article with be retracted from the journal. The second
would be used when an Letter to the Editor replies to an earlier published article,
and **[cito:updates]** when an Correction was published.

@@

This data in Wikidata can then be accessed via Scholia and Scholia tells use some
overall statistics of the number of annotations, which we reported on about a year
ago too [@citesAsDataSource:Willighagen_2021]. Since last year and written down
on August 25, the number of annotations and the number of annotated citations
have almost doubled (from 377 to 603 and from 304 to 494, respectively). The first
number is higher because one citation can have more than one citation intention.
To continue, the current number of citations are citing 387 articles in 141 different
scholarly journals, and they are found in 98 articles in 48 different journals
(see Figure 2) [@citesAsDataSource:scholiaCiTOStatistics].

It must be noted that the *Journal of Cheminformatics* is only one possible source,
but is still the only journal that uses CiTO annotation explicitly in the articles itself.
And with 335 annotated citations in 32 articles it also is the major source of
CiTO annotations in Wikidata at the time of writing.
However, CiTO intention annotations in Wikidata can come from other sources too
and be added both manually and automatically using the tools around Wikidata.
When all annotation is combined, Scholia shows us that **[cito:citesAsAuthority]**
is the most used intention, with 226 annotated citations (out of 603) in 38 articles.
**[cito:usesMethodIn]** follows with 102 annotated citations.

@@

In the first two years after the start of the Pilot, including the seminal editorial [@extends:Willighagen_2020],
the *Journal of Cheminformatics* has published fifteen articles with explicit CiTO annotation:
three Editorials, four Research articles, two Database, and one of each of Data Note, Software,
Letter to the Editor, Letter Response, Educational, and Methodology. Ten were published in
the first year (Table 1) and five in the second year (Table 2). Each article annotated
one or more citations with CiTO intentions, and several articles annotated every citation,
far exceeding the original anticipation.

@@

To make life easier for authors, and following a Twitter discussion in Spring 2021, a Markdown template was developed with native CiTO support: https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/markdown.html. Here, the author merely indicates the CiTO type when they cite the article. This is using a method introduced by Krewinkel et al. [@usesMethodIn:Krewinkel2017].
The manuscript can then be converted to a Microsoft Word file with Pandoc (https://pandoc.org/) for submission to the journal.

The *Journal of Cheminformatics* template is available from the journal's GitHub organization, and authors and editors should feel free to adapt it to the needs of other journals. The BioHackrXiv ([https://biohackrxiv.org/](https://biohackrxiv.org/))
preprint server also support CiTO annotations [@citesAsDataSource:Prins2021] and this template can be found at
[https://github.com/biohackrxiv/publication-template](https://github.com/biohackrxiv/publication-template).

@@

The general availability of open citation allows us to recover such more complex reuse scenarios
using the citation networks. Moreover, when the citations are annotated, we can zoom in on reuse
networks. Figure 3 shows a method reuse network for articles with explicit annotation in red. The
network shows a few article that use methods in two cited articles, e.g. *Kohulan Rajan, 2021*
and *Henning Otto Brinkhaus, 2022* [@Rajan2021,@Brinkhaus2022]. Of course, if we do not limit
the network to a subset of articles (here, *Journal of Cheminformatics* articles with CiTO
annotation), the network becomes more interesting, but also much more complex. Network analysis
approaches can then be used instead of network visualization.

@@

Because many publishing platforms currently do not support display of citation level intention annotation,
the default model only provides the annotation in the bibliography. The choice was made to use the `bibnotes`
idea which allows giving additional notes to a reference in the bibliography. The CiTO Pilot uses
the *safe CURIE* standard, compatible with the typesetting in the *Journal of Cheminformatics*.
This makes is easier to text mine the annotations by downstream tools.

This links to a limitation to the current use of CiTO annotation: the citation-level annotation may be
supported by some authoring systems (Markdown), even then the depiction may not be supported. If we convert
a Markdown file to a Microsoft Word file, the annotation is kept in the bibliography. However, when
writing a manuscript in Microsoft Word, LibreOffice, or Google Docs, combined with a reference manager
like Zotero, the CiTO annotation cannot be stored as part of the manuscript easily. The problem here is
that CiTO annotation in the reference manager are no longer linked to when they are cited. The workaround
is to add the CiTO annotation after completion of the Word document, directly to the bibliography.

LaTeX users may find them in a situation between that of Markdown users and reference manager users:
only if a manuscript-level Bib(La)TeX file is used the CiTO annotation can be added as notes to the
BiBTeX file. This is explained in this guidance document:
[https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/latex.html](https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/latex.html).

@@

With fourteen articles published in the CiTO Collection, the adoption is not as high as one would have hoped.
Nevertheless, the pilot is triggering interest with both authors and publishing platforms. The
*Journal of Cheminformatics* has already published a few more articles with CiTO annotation and
a search on the preprint servers ResearchSquare and ChemRxiv show a few more manuscripts. The support
by BioHackrXiv is a nice example of adoption beyond BMC.

The Scholia
use case shows how the information can be used downstream. Further uptake of this idea of typed citations
depends on the combined willingness of journal editors, authors, publishers and indexing services alike.
The rise of services like scite.ai shows that the research community is ready this kind of information.
Logical steps include support of distributing citation typing annotation via platforms like CrossRef
or EuropePMC.

Because all innovation requires a critical amount of adoption to be accepted, additional sources of
CiTO annotation will be welcome. For example, providing CiTO annotation via nanopublications would
allow collections of annotations to be shared. Then, when archived on Zenodo and therefore citable,
these annotations can be included in analysis as a trusted or at least citable source.

@@

CiTO annotation in the *Journal of Cheminformatics* are available from the journal's articles. CiTO
annotation data in Scholia is available from [Wikidata](https://wikidata.org/). Markdown templates that
support CiTO are available from
[https://github.com/jcheminform/markdown-jcheminf](https://github.com/jcheminform/markdown-jcheminf) and
[https://github.com/biohackrxiv/publication-template](https://github.com/biohackrxiv/publication-template).

>>>>

---
title: "Two years of explicit CiTO annotations"
subtitle: "Type: Correspondence"
author:
- Egon Willighagen [1,*,orcid:0000-0001-7542-0286]
link-citations: yes
geometry: margin=2.5cm
output:
  html_document:
    pandoc_args:
    - --lua-filter=filters/extract-cito.lua
    - --citeproc
    - --lua-filter=filters/insert-cito-in-ref.lua
  pdf_document:
    pandoc_args:
    - --lua-filter=filters/extract-cito.lua
    - --citeproc
    - --lua-filter=filters/insert-cito-in-ref.lua
  word_document:
    pandoc_args:
    - --lua-filter=filters/extract-cito.lua
    - --citeproc
    - --lua-filter=filters/insert-cito-in-ref.lua
# RStudio: cito-bibliography: bibliography.bib
# Other: bibliography: bibliography.bib
cito-bibliography: bibliography.bib
csl: journal-of-cheminformatics.csl
---

@@

Copyright (c) Egon Willighagen; CC-BY 4.0 International

@@

Communicating new research findings is still primarily done by written texts in the form of scholarly
articles, books, and book chapters. To not having to repeat past research by themselves or others,
authors cite relevant research [@agreesWith:citesAsAuthority:Shotton_2010]. However, the reasons why
authors cite literature vary, which complicates how we use citations [@agreesWith:citesAsAuthority:Nicholson_2021]. 
Typing citations is therefore of interest: it allows us to navigate literature more easily: it points
us to essential research methods, data, and can warn us of research that cannot be reproduced,
or others disagree with [@Nicholson_2021]. Indeed, it helps us understand the history of science
[@obtainsBackgroundFrom:Park2023].

@@

The pilot consisted of a couple of components and the editorial explains some of them [@Willighagen_2020].
The Citation Typing Ontology was selected to express the intention [@usesMethodIn:Shotton_2010],
the intention is expressed a compact identifier wrapped in square brackets, also called a
safe CURIE, standard proposed by the W3C in 2010 [@usesMethodIn:McCarron2010;@usesMethodIn:Wimalaratne_2018].
The *cito* prefix is registered in Bioregistry [@usesDataFrom:Hoyt2022].
The *bibnotes* concept of the Springer Nature publishing platform was used as carrier.
Authors are guided by a landing page consisting of a BMC Collection at
[https://www.biomedcentral.com/collections/cito](https://www.biomedcentral.com/collections/cito)
and author guidelines explaining to authors
how they can add the annotations with their favorite editor at
[https://jcheminform.github.io/jcheminform-author-guidelines/cito](https://jcheminform.github.io/jcheminform-author-guidelines/cito).

Because the CiTO ontology has many terms for many different citation intentions,
we made a selection of CiTO terms authors could use [@CiTOCollection]:
**[cito:citesAsDatasource]** to indicate
to a source the provides data to back up a claim, **[cito:usesDataFrom]** to indicate that the authors
reused data, **[cito:usesMethodIn]** when a method or protocol explain in that source is used,
and a few more general intentions like **[cito:discusses]**, **[cito:extends]**, **[cito:agreesWith]**,
and **[cito:disagreesWith]**.
The journal itself would adopt the following CiTO annotations: **[cito:retracts]**,
**[cito:repliesTo]**, and **[cito:updates]**. Fortunately, it has not been used yet, but
the first would be used if an article with be retracted from the journal. The second
would be used when an Letter to the Editor replies to an earlier published article,
and **[cito:updates]** when an Correction was published.

@@

This data in Wikidata can then be accessed in multiple ways, including REST APIs and
an SPARQL interface. The latter is used by Scholia to tell us some
overall statistics of the number of annotations, which we reported on about a year
ago too [@citesAsDataSource:Willighagen_2021]. Since last year and recorded
on August 25 2022, the number of annotations and the number of annotated citations
have almost doubled (from 377 to 603 and from 304 to 494, respectively). The first
number is higher because one citation can have more than one citation intention.
To continue, the current number of citations are citing 387 articles in 141 different
scholarly journals, and they are found in 98 articles in 48 different journals
(see Figure 2) [@citesAsDataSource:scholiaCiTOStatistics].

It must be noted that the *Journal of Cheminformatics* is only one possible source
of CiTO annotations. As far as the author knows, it is still the only journal that uses
CiTO annotation explicitly in the articles itself. 
And with 335 annotated citations in 32 articles it also is the major source of
CiTO annotations in Wikidata at the time of writing.
However, CiTO intention annotations in Wikidata can come from other sources too
and be added both manually and automatically using the tools around Wikidata.
When all annotation is combined, Scholia shows us that **[cito:citesAsAuthority]**
is the most used intention, with 226 annotated citations (out of 603) in 38 articles.
**[cito:usesMethodIn]** follows with 102 annotated citations.

@@

In the two years of the Pilot, including the seminal editorial [@extends:Willighagen_2020],
the *Journal of Cheminformatics* published fifteen articles with explicit CiTO annotation:
three Editorials, four Research articles, two Database, and one of each of Data Note, Software,
Letter to the Editor, Letter Response, Educational, and Methodology. Ten were published in
the first year (Table 1) and five in the second year (Table 2). Each article annotated
one or more citations with CiTO intentions, and several articles annotated every citation,
far exceeding the original anticipation.

@@

To make life easier for authors, and following a Twitter discussion in Spring 2021, a Markdown template was developed with native CiTO support:
[https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/markdown.html](https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/markdown.html).
Here, the author indicates the CiTO type when they cite the article. This is using a method introduced by Krewinkel et al. [@usesMethodIn:Krewinkel2017].
The manuscript can then be converted to a Microsoft Word file with Pandoc ([https://pandoc.org/](https://pandoc.org/)) for submission to the journal. For publishers it will be interesting to note that the Pandoc can be directly converted in
the *journal article tag suite* (JATS) format [@citesForInformation:Beck2011;@citesAsPotentialSolution:krewinkel2022jats].

The *Journal of Cheminformatics* template is available from the journal's GitHub organization, and authors and editors
should feel free to adapt it to the needs of other journals. The BioHackrXiv ([https://biohackrxiv.org/](https://biohackrxiv.org/))
preprint server also support CiTO annotations [@citesAsDataSource:citesAsPotentialSolution:Prins2021] and this template can be found at
[https://github.com/biohackrxiv/publication-template](https://github.com/biohackrxiv/publication-template).

@@

The general availability of open citation allows us to recover such more complex reuse scenarios
using the citation networks. Moreover, when the citations are annotated, we can zoom in on reuse
networks. Figure 3 shows a method reuse network for articles with explicit annotation in red. The
network shows a few article that use methods in two cited articles, e.g. *Kohulan Rajan, 2021*
and *Henning Otto Brinkhaus, 2022* [@Rajan2021;@Brinkhaus2022]. Of course, if we do not limit
the network to a subset of articles (here, *Journal of Cheminformatics* articles with CiTO
annotation), the network becomes more interesting, but also much more complex. Network analysis
approaches can then be used instead of network visualization.

@@

Because many publishing platforms currently do not support display of citation level intention annotation,
the simplest model only provides the annotation in the bibliography. The Pilot made this choice to
be able to use the `bibnotes` approach which allows giving additional notes to a reference in the bibliography. The CiTO Pilot uses
the *safe CURIE* standard, compatible with the typesetting in the *Journal of Cheminformatics*.
This makes is easier to text mine the annotations by downstream tools, from both the HTML and PDF as
well as the JATS versions of the article.

However, this links also to the limitation to the current use of CiTO annotation: the citation-level annotation may be
supported by some authoring systems (Markdown), even then the depiction may not be supported. If we convert
a Markdown file to a Microsoft Word file, the annotation is kept in the bibliography. However, when
writing a manuscript in Microsoft Word, LibreOffice, or Google Docs, combined with a reference manager
like Zotero, the CiTO annotation cannot be stored as part of the manuscript easily. The problem here is
that CiTO annotation in the reference manager are no longer linked to when they are cited. The workaround
is to add the CiTO annotation after completion of the Word document, directly to the bibliography.

LaTeX users may find them in a situation between that of Markdown users and reference manager users:
only if a manuscript-level Bib(La)TeX file is used the CiTO annotation can be added as notes to the
BiBTeX file. This way, the CiTO annotation is specific for this manuscript and each manuscript can use
different annotations. This approach is explained in this guidance document:
[https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/latex.html](https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/latex.html).

From a use case perspective, it is easy to see how this kind of annotation can be used. For example,
we here showed examples of reuse of work. via **[[cito:usesMethodIn]** and **[cito:usesDataFrom]**.
Second, scite.ai is a clear use case of **[cito:agreesWith]** and **[cito:disagreesWith]** annotation,
though it makes an good case how such citation intentions can be extracted with text mining instead.
However, other use cases still need development and adoption, which brings us to the question:
what is next?

@@

With fifteen articles published in the CiTO Collection, the pilot triggered interest from authors.
The *Journal of Cheminformatics* has already published a few more articles with CiTO annotation
after August 2022 and a search on the preprint servers ResearchSquare and ChemRxiv [@citesAsAuthority:Mudrak2022] show a
few more manuscripts. The support by BioHackrXiv [@citesAsAuthority:Prins2022] is a nice example of
adoption beyond BMC. Further citation intention annotations will come from literature studies where
citations networks are characterized. For example, Duca *et al.* used CiTO annotations to describe
the citation network to retracted COVID-19 articles [@citesAsRecommendedReading:Duca2022].

Further uptake of this idea of typed citations
depends on the combined willingness of journal editors, authors, publishers and indexing services alike.
The rise of services like scite.ai shows that the research community is ready this kind of information.
Logical steps include support of distributing citation typing annotation via platforms like PlumX,
Altmetric.com, CrossRef, or EuropePMC. Support of CiTO annotation in the JATS format seems an
important step forward.

Because all innovation requires a critical amount of adoption to be accepted, additional sources of
CiTO annotation will be welcome. For example, providing CiTO annotations via a standardized format
like a spreadsheet-like format or nanopublications would allow collections of annotations to be shared,
such as that by Duca *et al.*. Then, when archived on Zenodo and therefore citable,
these annotations can be included in analysis as a trusted or at least citable source.

@@

CiTO annotation in the *Journal of Cheminformatics* are available from the journal's articles. CiTO
annotation data in Scholia is available from [Wikidata](https://wikidata.org/). Markdown templates that
support CiTO are available from
[https://github.com/jcheminform/markdown-jcheminf](https://github.com/jcheminform/markdown-jcheminf) and
[https://github.com/biohackrxiv/publication-template](https://github.com/biohackrxiv/publication-template).
An archive of all CiTO annotations in Wikidata is available from [https://doi.org/10.5281/zenodo.7513573](https://doi.org/10.5281/zenodo.7513573).
